---
title: "Crime Mapping: Your first crime map"
description: "Quickly walk (or jog) through the process of creating a complete crime map in R."
output: 
  learnr::tutorial:
    progressive: true
    css: "css/tutorial_style.css"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)

# Copy files
if (!dir.exists("css")) dir.create("css")
walk(
  dir("../css/"), 
  ~ file.copy(str_glue("../css/{.}"), str_glue("css/{.}"), overwrite = TRUE)
)
```



## Introduction

In this tutorial we will use R to produce a simple crime map. To do this we will
skip over lots of the detail of how R works and what choices we should make in
creating maps. We will return to all these details in future sessions.

This tutorial includes short chunks of R code that you can run directly in this
window. We will use these chunks of code to walk through the code we need to
produce a map. To run the code in each chunk, just click the `Run Code` button
in the top-right corner of each chunk. If you want to re-run a chunk of code,
you can click the `Start Over` button in the top-left corner of each chunk.

When you click `Run Code`, the output of the chunk will appear below it (there
may be a delay of a few seconds for more-complicated chunks).

To try out using a code chunk, click `Run Code` below – you should see the
message 'Hello, world!' printed below the box.

```{r exercise0, exercise=TRUE}
message("Hello, world!")
```

`message()` is a *function* – a piece of R code that does something. The thing
that the `messages()` function does is simple: it prints a message on the 
screen.

Now click `Next Topic` below to start this tutorial.



## Loading crime data


### Loading packages

Before we can work with our data, we first load packages of functions for use in
the analysis. To load these packages, click `Run Code`. This will produce
various messages, all of which you can safely ignore for now.



```{r exercise1, exercise=TRUE, exercise.lines=5}
# Load the R packages we need to analyse this data
library("ggmap")
library("sf")
library("tidyverse")
```

<div class="box extra-detail">

<h5 id="loading-box1-title" class="box-title">What do these messages mean?</h5>

<div id="loading-box1" class="box-content">
For now you don't need to worry about these messages, but if you really want to
know what they mean …

Some R packages make use of external services. For example, the `ggmap` package
can make use of mapping services provided by Google. Google requires anyone
using their service to comply with their terms of service, so the `ggmap`
package prints a message when you load it to remind you of this.

Other packages make use of other apps and utilities on your computer. For
example, the `sf` package makes use of a piece of software call GDAL that is 
used for managing spatial data. So that you know which version of GDAL is being
used, `sf` prints a message telling you.

In general, R packages use start-up messages to remind you of information that 
is not likely to be critical to your work, but which it might be useful to know
at some point in the future.
</div>

</div>

<script>
$("#loading-box1-title").click(function () { $("#section-loading-box1").toggle("slow") })
</script>



### Loading data

The first task in creating any crime map is to obtain the crime and other data
necessary. In many cases preparing the data for analysis and mapping will be a 
substantial task, but in this case we are going to use some pre-prepared crime
data together with a pre-drawn street map (which we will ask R to download
automatically when it draws the final map).

The data we will use will be records of homicides in the Downtown neighbourhood
of Atlanta, Georgia, in 2019. We can load the homicide data using the
`read_csv()` function. Click the `Run Code` button to load the data.

```{r exercise2-setup, include=FALSE}
# Load the R packages we need to analyse this data
suppressPackageStartupMessages(library("tidyverse"))
```

```{r exercise2, exercise=TRUE}
# Download the data directly from a URL and store it as an object
homicides <- read_csv("https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv")
```

The `read_csv()` function prints a message showing the name of each column in
the data and the type of data (number, text etc.) in each column. Again, you can
ignore this message for now.

We have stored the results of the `read_csv()` function in an R *object* called
`homicides`. An object in R is anything that stores any type of data. There are
many types of objects, but for this tutorial we don't need to explore these in
any more detail. All you need to remember for now is that objects store data and 
functions do things.



### Viewing the data

To check the data has been loaded correctly, we can view the loaded data using
the `head()` function, which by default prints the first six rows of a dataset
stored in an object. Click `Run Code` to view the data.



```{r exercise3-setup, message=FALSE, warning=FALSE, include=FALSE}
# Load the R packages we need to analyse this data
suppressPackageStartupMessages(library("tidyverse"))

# Download the data directly from a URL and store it as an object
homicides <- read_csv(
  "https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv",
  col_types = cols(
    report_number = col_double(),
    label = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
)
```

```{r exercise3, exercise=TRUE}
# Display the data
head(homicides)
```



The data contain four columns: a unique identifier for a homicide, a label 
describing when and where that homicide occurred, and the longitude and latitude 
of the homicide location. Depending on the width of your screen, you may need to 
click on the '▸' symbol to view all the columns in the data. We can use this 
data to plot the homicides on a map.

<div class="notewell">

**In the code `head(homicides)`, there are no quote marks around the word
`homicides`.**

Almost all programming languages will interpret words differently depending on
whether they have quotes around them. In this case, if you type the code
`head(homicides)` then R will print the first few rows of the data stored in the
`homicides` object. 

On the other hand, if you type the code `head("homicides")` or
`head('homicides')`, R will interpret this as an instruction to print the first
few elements of the literal text 'homicides'. Since the text 'homicides' contains
only one element (more about that later), `head("homicides")` will just print
the word 'homicides'.

</div>


## Processing the data

Before we can plot the data on a map, we have to complete some pre-processing 
steps. Having to process data before being able to analyse or visualise it is 
common in all types of data analysis, but spatial analysis often involves
additional processing that takes account of the special features of spatial
data.



### Converting the data into a spatial format

Two data-processing tasks are needed to produce this map. The first is to 
convert the data into a *simple features* or SF object, which is a special type 
of R object that can be used by functions that process spatial data. We will 
cover the details of the `st_as_sf()` function that converts our data into into 
an SF object later on. Click `Run Code` to convert the data into SF format.



```{r exercise4-setup, message=FALSE, warning=FALSE, include=FALSE}
# Load the R packages we need to analyse this data
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("tidyverse"))

# Download the data directly from a URL and store it as an object
homicides <- read_csv(
  "https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv",
  col_types = cols(
    report_number = col_double(),
    label = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
)
```

```{r exercise4, exercise=TRUE}
# Convert the data to a simple features object, which we can use in functions 
# that work on spatial data
homicides <- st_as_sf(
  homicides, 
  coords = c("longitude", "latitude"), 
  crs = 4326, 
  remove = FALSE
)
```

When you clicked `Run Code` above, it looked like nothing happened. This is
because the results of the code are stored in the `homicides` object. Do you 
remember what R code to use to view the contents of this object? Type that code 
into the box below and click `Run Code` to view the data.

<div class="box notewell">

**If you cannot remember how to view the contents of an object, you can click on
the `Solution` button to get help.**

As you go through these tutorials, try to avoid using the `Solution` button 
unless you have tried to input the correct code yourself. You will learn much
more if you try to work out the answer by referring back to an earlier page in
the tutorial or to your own notes.

Don't worry about getting the answer wrong – nothing bad will happen if you run
the wrong code in this tutorial and you can have as many attempts as you like to
get the answer right.

</div>


```{r exercise5-setup, message=FALSE, warning=FALSE, include=FALSE}
# Load the R packages we need to analyse this data
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("tidyverse"))

# Download the data directly from a URL and store it as an object
homicides <- read_csv(
  "https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv",
  col_types = cols(
    report_number = col_double(),
    label = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
)

# Convert the data to a simple features object, which we can use in functions 
# that work on spatial data
homicides <- st_as_sf(
  homicides, 
  coords = c("longitude", "latitude"), 
  crs = 4326, 
  remove = FALSE
)
```

```{r exercise5, exercise=TRUE}

```

```{r exercise5-solution}
# To view an object in R, use the `head()` function. To view the contents of the
# homicides object, copy the next line into the box below and click Run Code.
head(homicides)
```

The data looks identical to before running the function `st_as_sf()`, except 
that there is now an extra column called `geometry`. This column is important
because lots of functions in R can use that column to analyse and map the data
in space.



### Defining the area covered by the map

The second data-processing task is to define the area that we want to be covered
by the map. In this case, we will define an area that covers the data and 500
metres in every direction around the data, meaning we can see the area 
surrounding the homicide locations.

Defining the area of the map relative to the data requires us to use several R
functions one after the other, connected using the characters `%>%` (which are
called the *pipe* in R – we will cover it later). One of the benefits of mapping
in R rather than [WYSIWYG](https://en.wikipedia.org/wiki/WYSIWYG) mapping
software is that you can connect all of these commands together and run them at
once, rather than having to open the relevant dialogue boxes or clicking the
right button for each command in turn.

You don't need to understand what each of these R functions does – we will cover
them all in a future tutorial.


```{r exercise6, exercise=TRUE, exercise.setup = "exercise5-setup", exercise.lines=13}
# Define the area included in the map, based on a 500m buffer around the data
map_boundary <- homicides %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_as_sf() %>% 
  st_transform(26967) %>% 
  st_buffer(500 * 3.3) %>% 
  st_transform(4326) %>% 
  st_bbox() %>% 
  as.numeric()

head(map_boundary)
```



This code produces an R object containing co-ordinates that represent the
lower-left and upper-right corners of our map. For now, we won't worry about how
to interpret these numbers. Now that we've completed the data processing, we can
go on to produce the map itself.



## Draw the map

We are now ready to produce our map of homicides in downtown Atlanta. So that
people viewing the map will understand where the homicides occurred, we will
plot the homicides on top of a base layer showing streets, parks and other 
geographic features obtained from an 
[online web mapping service](http://maps.stamen.com/). Click `Run Code` to 
create the map.



```{r exercise7-setup, message=FALSE, warning=FALSE, include=FALSE}
# Load the R packages we need to analyse this data
suppressPackageStartupMessages(library("ggmap"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("tidyverse"))

# Download the data directly from a URL and store it as an object
homicides <- read_csv(
  "https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/downtown_homicides.csv",
  col_types = cols(
    report_number = col_double(),
    label = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
)

# Convert the data to a simple features object, which we can use in functions 
# that work on spatial data
homicides <- st_as_sf(
  homicides, 
  coords = c("longitude", "latitude"), 
  crs = 4326, 
  remove = FALSE
)

# Define the area included in the map, based on a 500m buffer around the data
map_boundary <- homicides %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_as_sf() %>% 
  st_transform(26967) %>% 
  st_buffer(500 * 3.3) %>% 
  st_transform(4326) %>% 
  st_bbox() %>% 
  as.numeric()
```

```{r exercise7, exercise=TRUE, exercise.lines=13, fig.width=7, fig.asp=1}
suppressMessages(get_stamenmap(map_boundary, zoom = 14, maptype = "toner-lite")) %>% 
  ggmap() +
  geom_label(
    aes(longitude, latitude, label = label), data = homicides, size = 2.5,
    lineheight = 1, hjust = 1, nudge_x = -0.0005
  ) +
  geom_point(
    aes(longitude, latitude), data = homicides, colour = "white", 
    fill = "orangered1", size = 4, shape = 21
  ) +
  labs(title = "Homicides in Downtown Atlanta, 2019") +
  theme_nothing()
```

You can change the appearance of the map by changing various parts of the code
above and clicking `Run Code` again. For example, you can change the colour of
the points that mark the homicides by changing the code `fill = "orangered1"` to
`fill = "mediumblue"`, or change the base map to a different style by changing
the code `maptype = "toner-lite"` to `maptype = "terrain"`. Each time you change
part of the code, click `Run Code` to see what changes on the map.

Once you have finished experimenting with changing the appearance of the map,
click `Next Topic` below.



## Putting the code together

Now we have walked through the different parts of the code, we can create a map
from scratch in a single block of code. In this example, we will map homicides
from in Glenrose Heights neighbourhood of Atlanta, and a different style of base
map. Since the area covered by the map is derived from the data itself, the
extent of the map will update automatically. Click `Run Code`.


```{r exercise8, exercise=TRUE, exercise.lines=33, fig.width=7, fig.asp=1}
# Load the R packages we need to analyse this data
# `suppressPackageStartupMessages()` hides the start-up messages we saw earlier
suppressPackageStartupMessages(library("ggmap"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("tidyverse"))

# Download the data directly from a URL and store it as an object
# `suppressMessages()` hides the column-specification messages we saw earlier
homicides <- suppressMessages(read_csv("https://github.com/mpjashby/crimemapping/raw/main/inst/extdata/glenrose_heights_homicides.csv"))

# Convert the data to a simple features object, which we can use in functions 
# that work on spatial data
homicides <- st_as_sf(
  homicides, 
  coords = c("longitude", "latitude"), 
  crs = 4326, 
  remove = FALSE
)

# Define the area included in the map, based on a 500m buffer around the data
map_boundary <- homicides %>% st_bbox() %>% st_as_sfc() %>% st_as_sf() %>% 
  st_transform(26967) %>% st_buffer(500 * 3.3) %>% st_transform(4326) %>% 
  st_bbox() %>% as.numeric()

# Plot the map
suppressMessages(get_stamenmap(map_boundary, zoom = 14, maptype = "toner")) %>% 
  ggmap() +
  geom_label(
    aes(longitude, latitude, label = label), data = homicides, size = 2.5,
    lineheight = 1, hjust = 0, vjust = 1, nudge_x = 0.0005
  ) +
  geom_point(
    aes(longitude, latitude), data = homicides, colour = "white", 
    fill = "mediumblue", size = 4, shape = 21
  ) +
  labs(title = "Homicides in Glenrose Heights, 2019") +
  theme_nothing()
```



## In summary

<div class="box welldone">

**Well done – you have finished your first mapping tutorial.**

</div>

In this tutorial you have learned how to load data into R, prepare it for use in
making a map and then used it to make your first crime map of this course. 

The map we have produced in this tutorial is effective for showing the locations of just a few crimes, but is
too limited to show more complicated patterns or larger datasets. In the 
following tutorials, we will learn how to produce more sophisticated maps and
spatial analysis. We will also learn how each of the functions that we have used
in this tutorial work.
